\morus: Biases in the Stream

### 2018-05-10 morusAC_02_Preliminaries.tex
Section 2

The \MORUS family supports two internal state sizes: 640 and 1280 bits,
referred to as \MORUS[640] and \MORUS[1280], respectively.
Three parameter sets are recommended: \MORUS[640] supports 128-bit keys and \MORUS[1280] supports either 128-bit or 256-bit keys. The tag size is 128 bits or shorter. The designers strongly recommend using a 128-bit tag.

The internal state of \MORUS is composed of five $q$-bit \emph{registers} $S_i$, $i \in\{0,1,2,3,4\}$, where $q = 128$ for \MORUS[640] and $q=256$ for \MORUS[1280]. The internal state of \MORUS may be represented as $S_0\|S_1\|S_2\|S_3\|S_4$.
Registers are themselves divided into four $q/4$-bit \emph{words}.
Throughout the paper, we denote the word size by $w = q/4$, i.e., $w=32$ for \MORUS[640] and $w=64$ for \MORUS[1280].

@@

\StateUpdate{} is composed of 5 rounds with similar operations.
The additional input $m^t$ is used in rounds 2 to 5, but not in round 1.
Each round uses the word-wise rotation (left circular shift) operation,
denoted $\lll_w$ in the following and \texttt{Rotl\_xxx\_yy} in the design document.
It divides a $q$-bit register value into 4 words of $w = q/4$ bits, and performs a rotation on each $w$-bit word.
The word-wise rotation constants $b_i$ for round $i$ are defined in \Cref{Tbl:rcon}.
Additionally, each round uses rotations on a whole $q$-bit register by a multiple of the word size,
denoted $\lll$ in the following and \texttt{<<<} in the design document.
The register-wise rotation constants $b_i'$ are also listed in \Cref{Tbl:rcon}.
%\todo{rounds $0\dots4$ or $1\dots5$?}
\begin{table}%[!bht]
  \centering
  \caption{Rotation constants $b_i$ for $\lll_w$ and $b_i'$ for $\lll$ in round $i$ of \texorpdfstring{\MORUS}{MORUS}.} \label{Tbl:rcon}
  \setlength{\tabcolsep}{6pt}
  \begin{tabular}{@{}l@{\qquad}rrrrr rrrrr@{}}
    \toprule
    & \multicolumn{5}{@{}c@{~}}{Word-wise rotation $\lll_w$} & \multicolumn{5}{@{}c@{}}{Register-wise rotation $\lll$} \\
    \cmidrule(r){2-6}
    \cmidrule(l){7-11}
    & $b_0$  & $b_1$  & $b_2$  & $b_3$  & $b_4$  
    & $b_0'$ & $b_1'$ & $b_2'$ & $b_3'$ & $b_4'$  \\
    %& \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
    \midrule
    \MORUS[640]  &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
    \MORUS[1280] & 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\
    \bottomrule
  \end{tabular}
\end{table}

Let `$\cdot$' denote the bit-wise AND operation. $S^{t+1} \leftarrow \StateUpdate{}(S^t,m^t)$ is defined as follows:
\begin{align*}
\textrm{Round 1:} &&
S^{t+1}_0 &\leftarrow ( S^t_0 \oplus (S^t_1 \cdot S^t_2) \oplus S^t_3 )            \lll_w b_0, &
S^t_3 \leftarrow S^t_3 \lll b'_0.\\
\textrm{Round 2:} &&
S^{t+1}_1 &\leftarrow ( S^t_1 \oplus (S^t_2 \cdot S^t_3) \oplus S^t_4 \oplus m_i ) \lll_w b_1, &
S^t_4 \leftarrow S^t_4 \lll b'_1.\\
\textrm{Round 3:} &&
S^{t+1}_2 &\leftarrow ( S^t_2 \oplus (S^t_3 \cdot S^t_4) \oplus S^t_0 \oplus m_i ) \lll_w b_2, &
S^t_0 \leftarrow S^t_0 \lll b'_2.\\
\textrm{Round 4:} &&
S^{t+1}_3 &\leftarrow ( S^t_3 \oplus (S^t_4 \cdot S^t_0) \oplus S^t_1 \oplus m_i ) \lll_w b_3, &
S^t_1 \leftarrow S^t_1 \lll b'_3.\\
\textrm{Round 5:} &&
S^{t+1}_4 &\leftarrow ( S^t_4 \oplus (S^t_0 \cdot S^t_1) \oplus S^t_2 \oplus m_i ) \lll_w b_4, &
S^t_2 \leftarrow S^t_2 \lll b'_4.
\end{align*}

@@

\subsubsection{Encryption.}
At each encryption step, a 16-byte (resp. 32-byte) message block $M_t$ is used to update the state in \MORUS[640] (resp. \MORUS[1280]), and a ciphertext $C_t$ is produced.

If the last message block is not a full block, 0 bits are used to pad it to 128 and 256 bits for \MORUS[640] and \MORUS[1280], respectively, and the padded full block is used to update the state. But only the partial block is encrypted. Note that if the message length denoted by `${\it msglen}$' is 0, encryption is skipped.

In \MORUS[640], Let $u = \lceil \frac{{\it adlen}}{128} \rceil$ and $v = \lceil \frac{{\it msglen}}{128} \rceil$. The following is performed for $t=0, 1, \ldots, v-1$:
\begin{align*}
C^t &\leftarrow M^t \oplus S^{u+t}_0 \oplus (S^{u+t}_1 \lll 96) \oplus (S^{u+t}_2 \cdot S^{u+t}_3),\\
S^{u+t+1} &\leftarrow \StateUpdate{}(S^{u+t},M^t).
\end{align*}

For \MORUS[1280], $u = \lceil \frac{{\it adlen}}{256} \rceil$ and $v = \lceil \frac{{\it msglen}}{256} \rceil$. The ciphertext block $C^t$ is computed as $C^t \leftarrow M^t \oplus S^{u+t}_0 \oplus (S^{u+t}_1 \lll 192) \oplus (S^{u+t}_2 \cdot S^{u+t}_3)$.

\subsubsection{Finalization.}
The finalization phase generates the authentication tag $T$ using 10 more \StateUpdate{} steps. We only discuss the case where $T$ is not truncated. The associated data length and the message length are used to update the state as follows.
\begin{enumerate}
\item ${\it tmp} \leftarrow {\it adlen}\|{\it msglen}$ for \MORUS[640] and ${\it tmp} \leftarrow {\it adlen}\|{\it msglen}\|0^{128}$ for \MORUS[1280], where ${\it adlen}$ and ${\it msglen}$ are represented as 64-bit integers.
\item $S^{u+v}_4 \leftarrow S^{u+v}_4 \oplus S^{u+v}_0.$
\item For $t = u+v, u+v+1, \ldots, u+v+9,$ compute
%\begin{equation*}
$S^{t+1} \leftarrow \StateUpdate{} (S^t, {\it tmp})$.
%\end{equation*}
\item For \MORUS[640], $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 96) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$. For \MORUS[1280], $S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 192) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$ is computed and the least significant 128 bits are produced as a tag $T$.
\end{enumerate}

\subsubsection{Security.}
With a 128-bit tag, integrity is claimed up to 128 bits and confidentiality is claimed up to the number of bits of the key. See Table~\ref{Tbl/security} for summary.
\begin{table}[!htb]
\centering
\caption{Security goals of \MORUS.} \label{Tbl/security}
\begin{tabular}{ccc}\hline
                        & Confidentiality (bits) & Integrity (bits) \\ \hline
\MORUS[640-128]   & 128                    & 128              \\
\MORUS[1280-128]  & 128                    & 128              \\
\MORUS[1280-256]  & 256                    & 128              \\ \hline
\end{tabular}
\end{table}

@@

We define the \emph{bias} of an event $E$ as:
\[
\bias(E) \eqdef 2\Pr(E)-1.
\]
Note that the bias is sometimes defined as half the above value in the literature. We also define the \emph{weight} of an event as:
\[
\weight(E) \eqdef -\log|\bias(E)|
\]
where $\log()$ denotes logarithm in base 2. By the piling-up lemma, the bias (resp. weight) of an XOR of independent variables is equal to the product (resp. sum) of their individual biases (resp. weights) \cite{eurocryptMatsui93}.

We also recall the following notation from the previous section, where an \emph{encryption step} refers to one call to the \StateUpdate{} function:\\
\centerline{
\def\arraystretch{1.25}
\begin{tabular}{lcl}
$C^t$ &:& the ciphertext block output during the $t$-th encryption step.\\
$C^t_j$ &:& the $j$-th bit of $C^t$, with $C^t_0$ being the rightmost bit.\\
$S^t_i$ &:& the $i$-th register at the beginning of $t$-th encryption step.\\
$S^t_{i,j}$ &:& the $j$-th bit of $S^t_i$, with $S^t_{i,0}$ being the rightmost bit.
\end{tabular}
}
In the above notation, bit positions are always taken modulo the register size, i.e. 128 for \MORUS[640] and 256 for \MORUS[1280].

>>>>

The \MORUS family supports two internal state sizes: 640 and 1280 bits,
referred to as \MORUS[640] and \MORUS[1280], respectively.
Three parameter sets are recommended: \MORUS[640] supports 128-bit keys and \MORUS[1280] supports either 128-bit or 256-bit keys. The tag size is 128 bits or shorter. The designers strongly recommend using a 128-bit tag.
%
%\subsubsection{Security.}
With a 128-bit tag, integrity is claimed up to 128 bits and confidentiality is claimed up to the number of key bits (\Cref{Tbl/security}).

\begin{table}[b]
\caption{Security goals of \MORUS.} \label{Tbl/security}
\centering
\begin{tabular}{@{}l@{\quad}c@{\qquad}c@{}}\toprule
                        & Confidentiality (bits) & Integrity (bits) \\ \midrule
\MORUS[640-128]   & 128                    & 128              \\
\MORUS[1280-128]  & 128                    & 128              \\
\MORUS[1280-256]  & 256                    & 128              \\ \bottomrule
\end{tabular}
\end{table}

\subsubsection{State.}
The internal state of \MORUS is composed of five $q$-bit \emph{registers} $S_i$, $i \in\{0,1,2,3,4\}$, where $q = 128$ for \MORUS[640] and $q=256$ for \MORUS[1280]. The internal state of \MORUS may be represented as $S_0\|S_1\|S_2\|S_3\|S_4$.
Registers are themselves divided into four $q/4$-bit \emph{words}.
Throughout the paper, we denote the word size by $w = q/4$, i.e., $w=32$ for \MORUS[640] and $w=64$ for \MORUS[1280].

@@

\StateUpdate{} is composed of 5 rounds with similar operations.
The additional input $m^t$ is used in rounds 2 to 5, but not in round 1.
Each round uses the word-wise rotation (left circular shift) operation,
denoted $\lll_w$ in the following and \texttt{Rotl\_xxx\_yy} in the design document.
It divides a $q$-bit register value into 4 words of $w = q/4$ bits, and performs a rotation on each $w$-bit word.
The word-wise rotation constants $b_i$ for round $i$ are defined in \Cref{Tbl:rcon}.
Additionally, each round uses rotations on a whole $q$-bit register by a multiple of the word size,
denoted $\lll$ in the following and \texttt{<<<} in the design document.
The register-wise rotation constants $b_i'$ are also listed in \Cref{Tbl:rcon}.
%\todo{rounds $0\dots4$ or $1\dots5$?}

$S^{t+1} \leftarrow \StateUpdate{}(S^t,m^t)$ is defined as follows, where $\cdot$ denotes bit-wise AND, $\oplus$ is bit-wise XOR, and $m_i$ is defined depending on the context:
\begin{align*}
\textrm{Round 1:} &&
S^{t+1}_0 &\leftarrow ( S^t_0 \oplus (S^t_1 \cdot S^t_2) \oplus S^t_3 )            \lll_w b_0, &
S^t_3 \leftarrow S^t_3 \lll b'_0.\\
\textrm{Round 2:} &&
S^{t+1}_1 &\leftarrow ( S^t_1 \oplus (S^t_2 \cdot S^t_3) \oplus S^t_4 \oplus m_i ) \lll_w b_1, &
S^t_4 \leftarrow S^t_4 \lll b'_1.\\
\textrm{Round 3:} &&
S^{t+1}_2 &\leftarrow ( S^t_2 \oplus (S^t_3 \cdot S^t_4) \oplus S^t_0 \oplus m_i ) \lll_w b_2, &
S^t_0 \leftarrow S^t_0 \lll b'_2.\\
\textrm{Round 4:} &&
S^{t+1}_3 &\leftarrow ( S^t_3 \oplus (S^t_4 \cdot S^t_0) \oplus S^t_1 \oplus m_i ) \lll_w b_3, &
S^t_1 \leftarrow S^t_1 \lll b'_3.\\
\textrm{Round 5:} &&
S^{t+1}_4 &\leftarrow ( S^t_4 \oplus (S^t_0 \cdot S^t_1) \oplus S^t_2 \oplus m_i ) \lll_w b_4, &
S^t_2 \leftarrow S^t_2 \lll b'_4.
\end{align*}

\begin{table}[t]%[!bht]
  \centering
  \caption{Rotation constants $b_i$ for $\lll_w$ and $b_i'$ for $\lll$ in round $i$ of \texorpdfstring{\MORUS}{MORUS}.} \label{Tbl:rcon}
  \setlength{\tabcolsep}{6pt}
  \begin{tabular}{@{}l@{\qquad}rrrrr rrrrr@{}}
    \toprule
    & \multicolumn{5}{@{}c@{~}}{Word-wise rotation $\lll_w$} & \multicolumn{5}{@{}c@{}}{Register-wise rotation $\lll$} \\
    \cmidrule(r){2-6}
    \cmidrule(l){7-11}
    & $b_0$  & $b_1$  & $b_2$  & $b_3$  & $b_4$  
    & $b_0'$ & $b_1'$ & $b_2'$ & $b_3'$ & $b_4'$  \\
    %& \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
    \midrule
    \MORUS[640]  &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
    \MORUS[1280] & 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\subsubsection{Encryption.}
Next, the message is processed in blocks $M_t$ of  $q \in \{128, 256\}$ bits to update the state and produce the ciphertext blocks $C_t$.
If the last message block is not a full block, zero-bits are used to pad it to 128 or 256 bits for \MORUS[640] and \MORUS[1280], respectively, and the padded full block is used to update the state. But only the partial block is encrypted. Note that if the message length denoted by $|M|$ is 0, encryption is skipped.
Let $u = \lceil |A|/q \rceil$ and $v = \lceil |M|/q \rceil$. The following is performed for $t=0, 1, \ldots, v-1$:
\begin{align*}
C^t &\leftarrow M^t \oplus S^{u+t}_0 \oplus (S^{u+t}_1 \lll b_2') \oplus (S^{u+t}_2 \cdot S^{u+t}_3),\\
S^{u+t+1} &\leftarrow \StateUpdate{}(S^{u+t},M^t).
\end{align*}

\subsubsection{Finalization.}
The finalization phase generates the authentication tag $T$ using 10 more \StateUpdate{} steps. We only discuss the case where $T$ is not truncated. The associated data length and the message length are used to update the state:
\begin{enumerate}
\item $L \leftarrow |A| \mathrel\| |M|$ for \MORUS[640] or $L \leftarrow |A| \mathrel\| |M| \mathrel\|0^{128}$ for \MORUS[1280], where $|A|, |M|$ are represented as 64-bit integers.
\item $S^{u+v}_4 \leftarrow S^{u+v}_4 \oplus S^{u+v}_0.$
\item For $t = u+v, u+v+1, \ldots, u+v+9,$ compute
%\begin{equation*}
$S^{t+1} \leftarrow \StateUpdate{} (S^t, {\it tmp})$.
%\end{equation*}
%\item For \MORUS[640], $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 96) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$. For \MORUS[1280], $S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 192) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$ is computed and the least significant 128 bits are produced as a tag $T$.
\item $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll b_2') \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$, or the least significant 128 bits of this value in case of \MORUS[1280].
\end{enumerate}

@@

In the following, we use linear approximations \cite{eurocryptMatsui93}
that hold with probability $\Pr(E) = \frac{1}{2} + \varepsilon$, i.e., they are biased with bias $\varepsilon$.
The \emph{correlation} $\corr(E)$ of the approximation and its \emph{weight} $\weight(E)$ are defined as
\begin{align*}
\corr(E) &\eqdef 2\Pr(E)-1 = 2 \varepsilon\,, \\
%We also use the \emph{weight} of the approximation, defined as
\weight(E) &\eqdef -\log_2|\corr(E)|\,,
\end{align*}
where $\log_2()$ denotes logarithm in base 2. By the piling-up lemma, the correlation (resp. weight) of an XOR of independent variables is equal to the product (resp. sum) of their individual correlations (resp. weights) \cite{eurocryptMatsui93}.

We also recall the following notation from the previous section, where an \emph{encryption step} refers to one call to the \StateUpdate{} function:\\
\centerline{
\def\arraystretch{1.25}
\begin{tabular}{lcl}
$C^t$ &:& the ciphertext block output during the $t$-th encryption step.\\
$C^t_j$ &:& the $j$-th bit of $C^t$, with $C^t_0$ being the rightmost bit.\\
$S^t_i$ &:& the $i$-th register at the beginning of $t$-th encryption step.\\
$S^t_{i,j}$ &:& the $j$-th bit of $S^t_i$, with $S^t_{i,0}$ being the rightmost bit.
\end{tabular}
}
In the above notation, bit positions are always taken modulo the register size $q$, i.e., $q=128$ for \MORUS[640] and $q=256$ for \MORUS[1280].
